Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-02-08 Paper.Rmd
working on editing Q3 sections and accompanying sup mat material

Having demonstrated that non-N copula structure and asymmetric tail dependence are common
features of ecological data, we explored via mathematical 
models two possible mechanisms of this structure which 
seem likely to operate in diverse ecological scenarios. The first mechanism 
is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
It explores the hypothesis that non-N copula structure and asymmetric tail dependence 
between environmental variables measured through time 
in two places may bring about similarly 
structured dependence between ecological variables that are influenced by the 
environmental variables.

@@

<!--***DAN: note to self. Suppose deviations from the 
regression line for log met rate v log M are 
more costly at one end than at another, for whatever reason. That is a proximate
mechanism for asymmetric tail dependence-->

@@

# Results for Q2: Moran-like effects and asymmetric dependencies \label{Results_Q2}

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=12 cm]{./Results/asym_sens_results/a_0.2_b_0.1_r_0.8_asym_sens.pdf}
\caption{If environmental effects operate asymmetrically in their tails on
ecological variables, it can result in non-N copula structure and asymmetric 
tail dependence across space in the ecological variables. Shown are the last
$500$ points for (A) $g=g_1$ and (B) $g=g_2$ from simulations described in the text.\label{fig_ms_asym_b_0.1}}
\end{center}
\end{figure}

@@

<!--Question 3 analyses-->
# Concepts, methods and models for Q3: Consequences of non-normal copula structure and tail dependence for our understanding of biology \label{Consequences}

>>>>

Having demonstrated that non-N copula structure and asymmetric tail dependence are common
features of ecological data, we addressed Q2 by exploring via mathematical 
models two possible mechanisms of this structure which 
seem likely to operate in diverse ecological scenarios. The first mechanism 
is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
It explores the hypothesis that non-N copula structure and asymmetric tail dependence 
between environmental variables measured through time 
in two places may bring about similarly 
structured dependence between ecological variables that are influenced by the 
environmental variables.

@@

<!--***DAN: note to self. Suppose deviations from the 
regression line for log met rate v log M are 
more costly at one end than at another, for whatever reason. That is a proximate
mechanism for asymmetric tail dependence?-->

@@

# Results for Q2: Moran-like effects and asymmetric dependencies produce non-normal copula structure\label{Results_Q2}

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/asym_sens_results/a_0.2_b_0.1_r_0.8_asym_sens.pdf}
\caption{If environmental effects operate asymmetrically in their tails on
ecological variables, it can result in non-N copula structure and asymmetric 
tail dependence across space in the ecological variables. Shown are the last
$500$ points from simulations described in the text.\label{fig_ms_asym_b_0.1}}
\end{center}
\end{figure}

@@

<!--Question 3 analyses-->
# Background, concepts and methods for Q3: Consequences of non-normal copula structure for our understanding of ecology \label{Consequences}

We addressed Q3 using both models and data. The initial 
idea was presented in the Introduction that
the distribution (through time) of a spatially averaged quantity should depend on the 
nature of the dependence between the local quantities being averaged, including their
copula structure and tail dependence. For instance, suppose $E_i(t)$ are ecological
variables measured at locations $i=1,\ldots,N$ and times $t=1,\ldots,T$, and suppose 
the distribution and temporal dynamics of the spatial average $\sum_i E_i(t)/N$ is of
interest. The $E_i(t)$ could be, for instance, local abundances of a pest or 
exploited species, or local fluxes of a greenhouse gas. If  $E_i$ and $E_j$ are 
strongly right-tail dependent for all or most location pairs $i$ and $j$, then
exceptionally large values tend to occur at the same time in all or most locations.
This should tend to increase the skewness of the distribution through time of the
spatial average. Using similar reasoning, left-tail dependence between local variables
$E_i$ and $E_j$ will tend to decreases skewness of the spatial mean. Skewness of the 
spatial mean corresponds to the "spikiness" of the spatial-mean time series. 
The spatial-mean time series may be the quantity of principal importance for 
cases such as pest or resource abundance or CO$_2$ or methane flux.

We tested these ideas using the focal multivariate datasets of our earlier analyses:
green spruce aphid abundance data, leaf-curling plum aphid first flight data, \emph{Ceratium
furca} abundance data, and methane flux data (see Data and Table \ref{tab_multivar_summary}).
For each dataset we calculated the spatial mean, and then the skewness (through time) of that
mean. Then, for each dataset separately, we compared the value obtained 
to a distribution of values 
of the same quantity calculated for each of 10000 surrogate datasets that were 
produced from the empirical data by randomizing
it in a special way to have the copula structure of a multivariate-normal distribution
(Appendix \ref{SM-Surrog_multi}). The surrogate/randomized datasets had exactly the same
marginal distributions (i.e., distributions of values for each sampling location)
as the original data. 
They also had very similar Kendall or Spearman correlations (the algorithm had two versions,
one for preserving each correlation coefficient) between pairs of sampling locations
as the original data. But the surrogates had multivariate normal copula structure. 
Our comparisons therefore tested the null hypothesis that the skewness of the spatial 
mean took values on the empirical no different from what one would expect if the copula
structure of the data were multivariate normal, but the data were otherwise statistically
unchanged. Significant differences indicate that copula structure contributes to the skewness
of the spatial mean, i.e., to its "spikiness."

<!--***DAN: stopped here 2019 02 08 1106am central-->

### 2019-02-08 SupportingInformation.Rmd
working on editing Q3 sections and accompanying sup mat material

\newpage
# Consequence of non-Normal copula structure : Effect on Taylor's law \label{Consequence_TL_simulation}

>>>>

# Surrogate testing with multivariate datasets\label{Surrog_multi}

For multivariate datasets, surrogates were produced as follows:

- Given data $x_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$, let $\tau_{ij}$ be the Kendall
correlation of $x_i(t)$ and $x_j(t)$.  
- Calculate the covariance matrix of a multivariate normal distribution with standard-normal
marginals and with pairwise Kendall correlations equal to the $\tau_{ij}$. This is possible
using the `iTau` function of the `VineCopula` package.  
- Generate data $a_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$ by taking $T$ independent draws
from the the multivariate normal distribution of the previous step.  
- For any $i,t$ for which the datum $x_i(t)$ was missing, replace $a_i(t)$ by `NA`.
- For each $k$ and $i$ for which $a_i(t)$ is not `NA`, replace the 
$k$th-smallest (non-`NA`) element of the time series 
$a_i(t)$, $t=1,\ldots,T$ by the $k$th smallest (non-`NA`) 
element of $x_i(t)$, $t=1,\ldots,T$.
Call the result $b_i(t)$. This is the surrogate dataset.

Our code for this algorithm is `ncsurrog` in the `BIVAN` repository 
(\url{https://github.com/sghosh89/BIVAN}). Because, for each $i$, the final time series 
$b_i(t)$ is a permuted version of $x_i(t)$ (and with `NA`s for the same $t$), 
the marginal distributions of the surrogate dataset (i.e., the distributions of 
values for each sampling location, $i$) are exactly the same as for the original data.
Because ranks are preserved in passing from $a_i(t)$ to $b_i(t)$, pairwise Kendall 
correlations for the $b_i(t)$ will be the same as those of the $a_i(t)$. 
Discrepancies will come only from the presence of NAs and from sampling variation, 
and should be minor if the 
numbers of NAs are not too large and $T$ is not too small.

# Consequence of non-Normal copula structure : Effect on Taylor's law \label{Consequence_TL_simulation}
